Synthesizing Quotient Lenses

### 2018-03-08 papers/QuotientLenses/main.tex
More comments -- still not holding locks

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $c$ is semantically
equivalent to $\normalize{W(c)}{K(c)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(K(c))$. The {\em
normalize} combinator hence gives a sufficient condition for potential QREs to
be ``well-behaved'', particularly with respect to our strategy for synthesizing
quotient lenses which we shall introduce in Section~\ref{synth}.

In spite of this ``completeness'' property of the {\em normalize} combinator,
our experience with QREs shows that using the {\em project}, {\em squash} and
{\em perm} combinators as well as the regular operators applied to these is
much easier than using just {\em normalize} because of the compositional nature
of the other combinators.

@@

\section{QRE Lenses}
\label{QRE-lenses}
As we have seen, QREs express a broad class of equivalence relations
directly on regular languages.  QREs are therefore a good \textit{specification}
language for quotient lenses. In this section of the paper, we introduce
\textit{QRE Lenses}, a class of quotient lenses specified using QREs.

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, we define a {\em bijective quotient lens} $q : R /{\sim_R}
\Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$,
\begin{align*}
q.put(q.get(r)) &\sim_R r \text{, and}\\
q.get(q.put(s)) &\sim_R s
\end{align*}
Moreover, if $r \sim_R r'$, then $q.get(r) = q.get(r')$, and if $s \sim_S s'$,
then $q.put(s) = q.put(s')$.

In words, a bijective quotient lens $q$ from $R$ to $S$ modulo $\sim_R$
and $\sim_S$ is a pair of functions $q.get$ and $q.put$ such that $q.get$
respects $\sim_R$ and $q.put$ respects $\sim_S$, and such that the respective
functions defined on $\mathcal{L}(R)/{\sim_R}$ and $\mathcal{L}(S)/{\sim_S}$ are
mutual inverses of each other.

@@

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c_1$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c_1)$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c_1)
\end{align*}
\end{theorem}
\begin{proof}
We elide a full proof of this statement, since the proof is attained via a
straightforward induction on the derivation $q : c \Leftrightarrow c$ (the
full details can be found in the appendix {\sam{which I need to write}}). Since
bijective lenses are closed under strongly unambiguous regular operators, the
tricky case is composition. The composition case is more involved since we need
to show that we can get rid of the canonizers in the middle:

@@

\fi
\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c_1$ from the QREs $c, c_1$ and a set of input-output pairs
$\{(x_1, y_1),\allowbreak \ldots,\allowbreak (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c_1)$. We wish $q$ to map the equivalence class of $x_i$ to the
equivalence class of $y_i$ and vice versa:
\begin{align*}
q.get(x_i) &\equiv_{c_1} y_i \text{, and}\\
q.put(y_i) &\equiv_c x_i
\end{align*}

Our approach in synthesizing QRE lenses is guided by Theorem~\ref{normal
form}, which says that if there is a derivation $q :
c \Leftrightarrow c_1$ of a QRE lens, then there exists a bijective lens $\ell :
K(c) \Leftrightarrow K(c_1)$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c_1)
\end{align*}

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $c$ is semantically
equivalent to $\normalize{W(c)}{K(c)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(K(c))$. The {\em
normalize} combinator hence gives a sufficient condition for potential QREs to
be ``well-behaved'', particularly with respect to the strategy for synthesizing
quotient lenses which we shall introduce in Section~\ref{synth}.  \bcp{I
  think here we're mixing concerns of language design with the concerns of
  the underlying theory.  Indeed, I think it would be clearer to reorganize
  the whole section as follows: first present just normalize and discuss its
  properties; then observe that the side condition is annoying because it
  cannot be checked automatically; then introduce project, squash, and perm
  as a pragmatically motivated, extensible set of safe combinators that
  we've found useful for a broad range of examples.  This will also help
  address the complaint that extracting half of a lens is a hack.}

In spite of this ``completeness'' property of the {\em normalize} combinator,
our experience with QREs shows that using the {\em project}, {\em squash} and
{\em perm} combinators as well as the regular operators applied to these is
much easier than using just {\em normalize} because of the compositional nature
of the other combinators.  \bcp{Maybe this comment is not needed?  It seems
  obvious, besides the fact that these other combinators are safe while
  normalize requires a proof by hand.}

@@

\section{QRE Lenses}
\label{QRE-lenses}
As we have seen, QREs express a broad class of equivalence relations
directly on regular languages.  QREs are therefore a good \textit{specification}
language for quotient lenses. We next introduce
\textit{QRE Lenses}, a class of quotient lenses specified using QREs.

\bcp{A question about variable names: If we use $R$ and $S$ for plain
  regular expressions, it seems inconsistent to use $c$ for QREs.  (And even
  more inconsistent to use $c$ and $c'$ or $c$ and $c_1$!) I suggest $P$ and
  $Q$ for QREs (with $Q$ as the default when we are only talking about
  one).}

Given regular expressions $R, S$ and equivalence relations
$\sim_R$ and $\sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, we\bcp{can we clarify that this is exactly Foster et al's definition?} define a {\em bijective quotient lens} $q : R /{\sim_R}
\Leftrightarrow S/{\sim_S}$ to be a pair of functions $q.get :
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that for all $r \in \mathcal{L}(R)$ and $s
\in \mathcal{L}(S)$,
\begin{align*}
q.put(q.get(r)) &\sim_R r \text{, and}\\
q.get(q.put(s)) &\sim_R s
\end{align*}
Moreover, if $r \sim_R r'$, then $q.get(r) = q.get(r')$, and if $s \sim_S s'$,
then $q.put(s) = q.put(s')$.  \bcp{We should emphasize that this is
  precisely the desired loosening of the too-strict plain lens laws.}

In words, a bijective quotient lens $q$ from $R$ to $S$ modulo $\sim_R$
and $\sim_S$ is a pair of functions $q.get$ and $q.put$ such that $q.get$
respects $\sim_R$ and $q.put$ respects $\sim_S$, and such that the $get$ and
$put$ functions lifted to $\mathcal{L}(R)/{\sim_R}$ and
$\mathcal{L}(S)/{\sim_S}$ are mutual inverses.

\bcp{Typesetting: $q.get$ and $q.\get$ are subtly different, as are
  $q.put$ and $q.\lput$, and the latter forms are the correct ones.
  In general, you should never put a whole word in math mode without
  wrapping it in mathit: the inter-letter spacing will be wrong.}

@@

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c_1$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c_1)$ such that:
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c_1)
\end{align*}
\end{theorem}
\begin{proof}
We elide a full proof of this statement, since the proof is attained via a
straightforward induction on the derivation $q : c \Leftrightarrow c$ (the
full details can be found in the appendix {\sam{which I need to write}}). Since
bijective lenses are closed under strongly unambiguous regular operators, the
tricky case is composition. The composition case is more involved since we need
to show that we can get rid of the canonizers in the middle:

@@

\fi
\section{Synthesizing QRE Lenses}
\label{synth}
We next demonstrate how to use previous results from~\cite{optician} to
synthesize a quotient lens 
$q: c \Leftrightarrow c_1$ from the QREs $c, c_1$ and a set of input-output pairs
$\{(x_1, y_1),\allowbreak \ldots,\allowbreak (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c_1)$. We wish $q$ to map the equivalence class of $x_i$ to the
equivalence class of $y_i$ and vice versa:
\begin{align*}
q.get(x_i) &\equiv_{c_1} y_i \text{, and}\\
q.put(y_i) &\equiv_c x_i
\end{align*}

\bcp{Another question about choice of variables: if we rename $c$ and $c_1$
  to $P$ and $Q$, could we also choose names like $p$ and $q$ for their
  elements?  (I guess $q$ was used to range over QRE-lenses, so we can't use
  that suggestion without further renaming...)}

Our approach in synthesizing QRE lenses is guided by Theorem~\ref{normal
form}, which says that, if there is a derivation $q :
c \Leftrightarrow c_1$ of a QRE lens, then there exists a bijective lens $\ell :
K(c) \Leftrightarrow K(c_1)$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c_1).
\end{align*}
